# Course Project in the Coursera Data Science Specialization 'Regression Models' - Analysis of the 'mtcars' data set.

---
title: "RegMod_CP_MTCARS.Rmd", author: "Mark A. Jack", date: "July 25, 2015", output: pdf_document
---

## Executive Summary
In this course project, data set 'mtcars' is analyzed with regards to the two questions whether an automatic or manual transmission is better for a car's gas mileage (miles per gallon) and to quantify the mpg difference between automatic and manual transmissions. The column 'mpg' in the 'mtcars' data set is selected as outcome. The following predictor variables are chosen: 'factor(am)' as a two-level factor variable ('am=0' for manual and 'am=1' for automatic transmission); 'wt' measuring a car's weight in units of 1000 lbs; 'hp' measuring a car's horse power in units hp; and 'factor(am)*wt' as interaction term between variables 'am' and 'wt'. Separate linear regression fits are conducted for a comparison of 'mpg' as outcome and 'factor(am)' and 'wt' as predictors and 'factor(am) wt' as interaction term and a second comparison with 'mpg' and 'factor(am)' and 'hp' as outcome and predictor variables, respectively. A general decrease of miles per gallon can be observed with an increase in horse powers, as expected which is independent of the selected group, i.e. whether automatic (am=1) or manual transmission (am=0) is being assumed. An average downward shift of about 5 miles per gallon in mpg can be seen when switching from 'am=1' to 'am=0'. In the other comparison, separate linear fits occur with different slopes for the two groups 'am=1' and 'am=0' when 'mpg' is fitted versus 'wt'. This differentiation is due to a relevant interaction term 'factor(am) wt'. Automatic transmission (am=1) appears favorable for higher gas mileage when weights are smaller while for increased weights a cross-over happens where manual transmission (am=0) starts yielding higher mpg values. This crossover happens at a weight of around 3000 lbs.  
```{r, include=FALSE, cache=FALSE}
# R Libraries 'plyr', 'dplyr', 'Hmisc', 'reshape2', 'ggplot2', and 'knitr' are loaded for this data analysis.
library(plyr)
library(dplyr)
library(Hmisc)
library(reshape2)
library(ggplot2)
library(knitr)
```
## Data Processing
The data set 'mtcars' includes the data of of 32 vehicle and 10 column variables. We calculate the mean, median, standard deviation, minimum and maximum for each of the three key continuous variables in this analysis, 'mpg', 'wt', and 'hp', for each group 'manual transmission (am=0)' and 'automatic transmission (am=1)'. The quick analysis of the data highlights a trend of larger mpg values for automatic transmission (am=1) versus manual transmission (am=0), i.e. on average 17 mpg for 'am=0' and 24 mpg for 'am=1'. However, this also needs to be seen in the context of larger average weight of vehicles of 3769 lbs (am=0) versus 2411 lbs (am=1), respectively, while a vehicle's horsepower shows a smaller average of 126 hp for 'am=1' versus 160 hp for 'am=0' while there is also a larger spread of 'hp' values in the first case.
```{r, include=FALSE, cache=FALSE}
data(mtcars)
dim(mtcars)
head(mtcars, 1)
#
mtcars_melt <- melt(mtcars, id.vars = "am", measure.vars = "mpg", na.rm = TRUE)
mtcars_select <- group_by(mtcars_melt, am, variable)                       
mtcars_mean <- ddply(mtcars_select, c("am"), summarise, mean = mean(value, na.rm = TRUE))
mtcars_median <- ddply(mtcars_select, c("am"), summarise, median = median(value, na.rm = TRUE))
mtcars_sd <- ddply(mtcars_select, c("am"), summarise, sd = sd(value, na.rm = TRUE))
mtcars_min <- ddply(mtcars_select, c("am"), summarise, min = min(value, na.rm = TRUE))
mtcars_max <- ddply(mtcars_select, c("am"), summarise, max = max(value, na.rm = TRUE))
mtcars_stat <- mtcars_mean
mtcars_stat$median <- mtcars_median$median
mtcars_stat$sd <- mtcars_sd$sd 
mtcars_stat$min <- mtcars_min$min 
mtcars_stat$max <- mtcars_max$max
#
mtcars_melt_wt   <- melt(mtcars, id.vars = "am", measure.vars = "wt", na.rm = TRUE)
mtcars_select_wt <- group_by(mtcars_melt_wt, am, variable)                       
mtcars_mean_wt   <- ddply(mtcars_select_wt, c("am"), summarise, mean = mean(value, na.rm = TRUE))
mtcars_median_wt <- ddply(mtcars_select_wt, c("am"), summarise, median = median(value, na.rm = TRUE))
mtcars_sd_wt   <- ddply(mtcars_select_wt, c("am"), summarise, sd = sd(value, na.rm = TRUE))
mtcars_min_wt  <- ddply(mtcars_select_wt, c("am"), summarise, min = min(value, na.rm = TRUE))
mtcars_max_wt  <- ddply(mtcars_select_wt, c("am"), summarise, max = max(value, na.rm = TRUE))
mtcars_stat_wt <- mtcars_mean_wt
mtcars_stat_wt$median <- mtcars_median_wt$median
mtcars_stat_wt$sd  <- mtcars_sd_wt$sd 
mtcars_stat_wt$min <- mtcars_min_wt$min 
mtcars_stat_wt$max <- mtcars_max_wt$max
#
mtcars_melt_hp  <- melt(mtcars, id.vars = "am", measure.vars = "hp", na.rm = TRUE)
mtcars_select_hp <- group_by(mtcars_melt_hp, am, variable)                       
mtcars_mean_hp   <- ddply(mtcars_select_hp, c("am"), summarise, mean = mean(value, na.rm = TRUE))
mtcars_median_hp <- ddply(mtcars_select_hp, c("am"), summarise, median = median(value, na.rm = TRUE))
mtcars_sd_hp   <- ddply(mtcars_select_hp, c("am"), summarise, sd = sd(value, na.rm = TRUE))
mtcars_min_hp  <- ddply(mtcars_select_hp, c("am"), summarise, min = min(value, na.rm = TRUE))
mtcars_max_hp  <- ddply(mtcars_select_hp, c("am"), summarise, max = max(value, na.rm = TRUE))
mtcars_stat_hp <- mtcars_mean_hp
mtcars_stat_hp$median <- mtcars_median_hp$median
mtcars_stat_hp$sd  <- mtcars_sd_hp$sd 
mtcars_stat_hp$min <- mtcars_min_hp$min 
mtcars_stat_hp$max <- mtcars_max_hp$max
```
```{r, include=FALSE, cache=FALSE}
head(mtcars_stat)
head(mtcars_stat_wt)
head(mtcars_stat_hp)
```
## Linear Regression Analysis
In the selection of proper variables for this analysis, we examine the p-values of the linear regression fits of outcome 'mpg' to the final selection of continuous predictor variables 'wt' and 'hp' together with the factor variable 'factor(am)'. This is highlighted in following model likelihood fits using the R tool 'anova' where all linear regression models as described above up are listed. Each of the above variables are highly significant (p-values less than 0.05 for 95%-confidence level). While the interaction term 'factor(am) wt' also proves to be highly significant, the interaction term 'factor(am) hp' ends up being omitted for the final analysis.
```{r, include=FALSE, cache=FALSE}
lm_mtcars <- lm(data = mtcars, mpg ~ factor(am))
#summary(lm(data = mtcars, mpg ~ factor(am)))$coef
#
lm_mtcars_wt <- lm(data = mtcars, mpg ~ factor(am) + wt)
#summary(lm(data = mtcars, mpg ~ factor(am) + wt))$coef
#
lm_mtcars_wt_hp <- lm(data = mtcars, mpg ~ factor(am) + wt + hp)
#summary(lm(data = mtcars, mpg ~ factor(am) + wt + hp))$coef
#
lm_mtcars_wt_hp_am <- lm(data = mtcars, mpg ~ factor(am) + wt + hp + factor(am)*wt)
#summary(lm(data = mtcars, mpg ~ factor(am) + wt + hp + factor(am)*wt))$coef
#
lm_mtcars_wt_hp_am2 <- lm(data = mtcars, mpg ~ factor(am) + wt + hp + factor(am)*wt + factor(am)*hp)
summary(lm(data = mtcars, mpg ~ factor(am) + wt + hp + factor(am)*wt + factor(am)*hp))$coef
```
```{r, echo=FALSE}
anova(lm_mtcars, lm_mtcars_wt, lm_mtcars_wt_hp, lm_mtcars_wt_hp_am, lm_mtcars_wt_hp_am2)
```
Thus, for the two linear regression fits of 'mpg' as outcome versus 'factor(am)', 'wt' and 'factor(am) wt' as predictor variables in the first fit and 'factor(am)' and 'hp' in the second fit we compute the following regression coefficients:
```{r, echo=FALSE}
lm_mtcars_wt_am <- lm(data = mtcars, mpg ~ factor(am) + wt + factor(am)*wt)
summary(lm(data = mtcars, mpg ~ factor(am) + wt + factor(am)*wt))$coef
#
lm_mtcars_hp_am <- lm(data = mtcars, mpg ~ factor(am) + hp)
summary(lm(data = mtcars, mpg ~ factor(am) + hp))$coef
```
The 95%-confidence level error bars for the regression coefficients are listed below:
```{r, echo=FALSE}
confint(lm_mtcars_wt_am)
confint(lm_mtcars_hp_am)
```
The Appendix summarizes the above findings and analysis in a series of different plots:

- Figure 1: Two histograms are shown to demonstrate the distribution of mpg values in the data set for the two groups 'automatic transmission (am=1)' and 'manual transmission (am=0)'.

- Figure 2: The data points for the two regression models are highlighted in red for group 'am=0 (manual)' and in black for 'am=1 (automatic)'. The left-hand plot shows two linear regression fits with different slopes with 'mpg' as outcome versus 'wt' as predictor (black and red lines) due to the included interaction term, 'factor(am) wt'. The right-hand plot shows two linear regression lines with equal slope when fitting 'mpg' versus 'hp'. An overall shift of average mpg values occurs for automatic versus manual transmission at constant hp values.

- Figures 3: Figure 3 shows the distribution of normalized residuals for each of the above two regression models. Normalization is done with respect to the standard deviation as computed assuming a t-student distribution due to the small sample size (only 32 vehicle brands compared). 

- Figure 4: These two plots compare residual values (y-axis) with fitted values (x-axis) of each model. A clear separation of the data of the two groups 'am=0' and 'am=1' can be seen towards lower and higher mpg values, respectively. 

## Appendix

### Plots:

#### Figure 1:
```{r, echo=FALSE}
par(mfrow = c(1,2))
par(mar = c(4,4,2,2))
facet_names <- list("0" = "Automatic", "1" = "Manual")
facet_labeller <- function(am, value){
  value <- as.character(value)
  return(facet_names[value])
}
g1 <- ggplot(mtcars, aes(mpg)) + geom_histogram(binwidth = 1, col = "blue")
g2 <- g1 + facet_grid(.~ am, labeller = facet_labeller) 
g3 <- g2 + labs(title = "Miles per gallon for automatic or manual transmission") 
g4 <- g3 + labs(x = "Miles per gallon [mpg]", y = "frequency")
g4
#dev.copy(png, file = "/Users/markjack/data/plot_mtcars_mpg.png")
#dev.off()
```

#### Figure 1.

#### Figure 2:
```{r, echo=FALSE}
par(mfrow = c(1,2))
plot(mtcars$wt, mtcars$mpg, pch=19, xlab = "weight [in 1000 lbs]", ylab = "miles per gallon [mpg]", 
     bg = "lightblue", col = "black", cex = 2, frame = FALSE)
points(mtcars$wt, mtcars$mpg, pch=19, col=((mtcars$am==0)*1+1))
abline(c(lm_mtcars_wt_am$coef[1],lm_mtcars_wt_am$coef[3]), col="red", lwd=3)
abline(c(lm_mtcars_wt_am$coef[1]+lm_mtcars_wt_am$coef[2],lm_mtcars_wt_am$coef[3]+lm_mtcars_wt_am$coef[4]), 
       col="black", lwd=3)
legend('topright', c("manual","autom."), pch=19, col=c("red", "black"), inset = .05)
#
plot(mtcars$hp, mtcars$mpg, pch=19, xlab = "horse power [hp]", ylab = "miles per gallon [mpg]", 
     bg = "lightblue", col = "black", cex = 2, frame = FALSE)
points(mtcars$hp, mtcars$mpg, pch=19, col=((mtcars$am==0)*1+1))
abline(c(lm_mtcars_hp_am$coef[1],lm_mtcars_hp_am$coef[3]), col="red", lwd=3)
abline(c(lm_mtcars_hp_am$coef[1]+lm_mtcars_hp_am$coef[2],lm_mtcars_hp_am$coef[3]), col="black", lwd=3)
legend('topright', c("manual","autom."), pch=19, col=c("red", "black"), inset = .05)
#
#dev.copy(png, file = "/Users/markjack/data/plot_lm_mpg_hp_hist.png")
#dev.off()
```

#### Figure 2.

```{r, include=FALSE, cache=FALSE}
e_wt_am <- rstudent(lm_mtcars_wt_am)
e_hp_am <- rstudent(lm_mtcars_hp_am)
fit_wt_am <- predict(lm_mtcars_wt_am)
fit_hp_am <- predict(lm_mtcars_hp_am)
```

#### Figure 3:
```{r, echo=FALSE}
par(mfrow = c(1,2))
#
plot(mtcars$wt, e_wt_am, xlab = "weight [in 1000 lbs]", ylab = "norm. residuals [t-dist.]", bg = "lightblue", col = "black", cex = 2, pch = 21, frame = FALSE)
abline(h=0, lwd=2)
#
for (i in 1 : n)
  lines(c(mtcars$wt[i], mtcars$wt[i]), c(e_wt_am[i], 0), col = "red", lwd = 2)
#
plot(mtcars$hp, e_hp_am, xlab = "horse power [hp]", ylab = "norm. residuals [t-dist.]", bg = "lightblue", col = "black", cex = 2, pch = 21, frame = FALSE)
abline(h=0, lwd=2)
#
for (i in 1 : n)
  lines(c(mtcars$hp[i], mtcars$hp[i]), c(e_hp_am[i], 0), col = "red", lwd = 2)
#dev.copy(png, file = "/Users/markjack/data/plot_resid_mpg_hp.png")
#dev.off()
```

#### Figure 3.

#### Figure 4:
```{r, echo=FALSE}
par(mfrow = c(1,2))
plot(fit_wt_am, e_wt_am, pch=19, xlab = "miles per gallon [predicted for 'mpg~wt']", 
     ylab = "normalized residuals [t-dist.]", bg = "lightblue", 
     col = "black", cex = 2, frame = FALSE)
points(fit_wt_am, e_wt_am, pch=19, col=((mtcars$am==0)*1+1))
legend('topleft', c("manual","autom."), pch=19, col=c("red", "black"), inset = .05)
#
plot(fit_hp_am, e_hp_am, pch=19, xlab = "miles per gallon [predicted for 'mpg~hp']", 
     ylab = "normalized residuals [t-dist.]", bg = "lightblue", 
     col = "black", cex = 2, frame = FALSE)
points(fit_hp_am, e_hp_am, pch=19, col=((mtcars$am==0)*1+1))
legend('topleft', c("manual","autom."), pch=19, col=c("red", "black"), inset = .05)
#
#dev.copy(png, file = "/Users/markjack/data/plot_fit_res_mpg_hp_hist.png")
#dev.off()
```

#### Figure 4.